---
title: "polishing benchmark"
format: html
editor: visual
---

# libraries

```{r}
#| echo: false
library(tidyverse)
library(lubridate)
library(flextable)
library(scales)
library(officer)
library(patchwork)
library(ggsci)
```

# functions

```{r}
read_trace_files <- function(path){
  # path <- trace_file_paths[2]
  config <- str_split(path, "/", simplify = TRUE)[,7]
  polishing <- str_extract(path, "reference|poa")
  mode <- str_extract(path, "cpu|gpu")
  cluster_per_polishing_file <- as.numeric(str_split(config, "_", simplify = TRUE)[,5])
  max_reads_per_cluster <- as.numeric(str_split(config, "_", simplify = TRUE)[,6])
  
  trace <-
    read_tsv(path, col_types = "iciccicccccccc") %>% 
    separate(name, c("pipeline", "workflow", "process", "label"), sep = ":| ", fill = "right") %>%
    select(
      pipeline, 
      workflow, 
      process,
      label,
      status,
      duration,
      realtime,
      peak_vmem, 
      peak_rss,
      submit
    ) %>%
    mutate(
      config = config,
      polishing = polishing,
      mode = mode,
      cluster_per_polishing_file = cluster_per_polishing_file,
      max_reads_per_cluster = max_reads_per_cluster,
      submit = as_datetime(submit)
    )
}
read_mutserve_files <- function(path){
  # path <- trace_file_paths[2]
  config <- str_split(path, "/", simplify = TRUE)[,7]
  polishing <- str_extract(path, "reference|poa")
  mode <- str_extract(path, "cpu|gpu")
  cluster_per_polishing_file <- as.numeric(str_split(config, "_", simplify = TRUE)[,5])
  max_reads_per_cluster <- as.numeric(str_split(config, "_", simplify = TRUE)[,6])
  
  trace <-
    read_tsv(path, show_col_types = FALSE) %>% 
    mutate(
      config = config,
      polishing = polishing,
      mode = mode,
      cluster_per_polishing_file = cluster_per_polishing_file,
      max_reads_per_cluster = max_reads_per_cluster
    )
}

read_cluster_sizes <- function(path){
  config <- str_split(path, "/", simplify = TRUE)[,6]
  polishing <- str_extract(path, "reference|poa")
  cluster_per_polishing_file <- as.numeric(str_split(config, "_", simplify = TRUE)[,5])
  max_reads_per_cluster <- as.numeric(str_split(config, "_", simplify = TRUE)[,6])
  
  read_tsv(path, show_col_types = FALSE) %>% 
    mutate(
      config = config, 
      polishing = polishing, 
      cluster_per_polishing_file = cluster_per_polishing_file, 
      max_reads_per_cluster = max_reads_per_cluster, 
      filenumber = as.numeric(str_split(log_file, "_|\\.", simplify = TRUE)[,4])
    )
}
convert_to_mb <- function(x) {
  x <- str_replace_all(x, ",", "") # remove commas
  x <- str_trim(x)                 # remove whitespace
  as.numeric(case_when(
    str_detect(x, "GB") ~ as.numeric(str_remove(x, " GB")),
    str_detect(x, "MB") ~ as.numeric(str_remove(x, " MB")) / 1024,
    str_detect(x, "KB") ~ as.numeric(str_remove(x, " KB")) / 1024^2,
    TRUE ~ 0
  ))
}
```

# paths

```{r}
trace_file_paths <- list.files(
  path = "../benchmarking/out/nextflow_stats/", 
  pattern = "trace.txt", 
  recursive = TRUE, 
  full.names = TRUE
)

mutserve_file_paths <- list.files(
  path = "../benchmarking/out/mutserve/", 
  pattern = "consensus_parsed.txt", 
  recursive = TRUE, 
  full.names = TRUE
)

glued_cluster_sizes_paths <- list.files(
  path = "../benchmarking/out/glued_cluster_sizes/", 
  pattern = "glued_cluster_sizes.tsv", 
  recursive = TRUE, 
  full.names = TRUE
)
```

# load files

```{r}
trace_files <- 
  lapply(trace_file_paths, read_trace_files) %>% 
  bind_rows()

mutserve_files <- 
  lapply(mutserve_file_paths, read_mutserve_files) %>% 
  bind_rows()

glued_cluster_sizes <- 
  lapply(glued_cluster_sizes_paths, read_cluster_sizes) %>% 
  bind_rows()

```

# tables

## trace files

```{r}
trace_files <- 
  trace_files %>% 
  mutate(
    mode = ifelse(mode == "cpu", "CPU", "GPU"), 
    polishing = ifelse(polishing == "reference", "Reference-based", "POA-based"), 
    process_order = case_when(
      str_detect(process, "ALIGN_CLUSTER") ~ 1,
      str_detect(process, "PARSE_BAM") ~ 2,
      str_detect(process, "CREATE_CONSENSUS") ~ 3,
      str_detect(process, "STITCH_CONSENSUS") ~ 4,
      str_detect(process, "POLISH_CLUSTER") ~ 1
    )
  )

trace_files_filtered <- 
  trace_files %>% 
  filter(str_detect(workflow, "UMI_POLISHING")) %>%
  filter(process %in% c("ALIGN_CLUSTER", "PARSE_BAM", "CREATE_CONSENSUS", "STITCH_CONSENSUS", "POLISH_CLUSTER")) %>% 
  filter(max_reads_per_cluster <= 50) %>% 
  filter(cluster_per_polishing_file <= 50)

trace_files_parsed_milliseconds <- 
  trace_files_filtered %>% 
  filter(str_detect(duration, "ms")) %>% 
  mutate(
    duration = as.duration(paste0(as.numeric(str_remove(duration, "ms")) / 1000, "s")),
    realtime = as.duration(paste0(as.numeric(str_remove(realtime, "ms")) / 1000, "s")), 
    peak_rss_mb = convert_to_mb(peak_rss)
  )

trace_files_parsed_minutes <- 
  trace_files_filtered %>% 
  filter(str_detect(duration, "ms", negate = TRUE)) %>% 
  mutate(
    duration = as.duration(str_replace(duration, "m |m$", "min ")),
    realtime = as.duration(str_replace(realtime, "m |m$", "min ")),
    cluster_per_polishing_file = as.numeric(cluster_per_polishing_file), 
    max_reads_per_cluster = as.numeric(max_reads_per_cluster), 
    peak_rss_mb = convert_to_mb(peak_rss)
  )

trace_files_parsed <- 
  trace_files_parsed_milliseconds %>% 
  bind_rows(trace_files_parsed_minutes)

duration_summary_per_process <- 
  trace_files_parsed %>% 
  group_by(polishing, cluster_per_polishing_file, mode, max_reads_per_cluster, process) %>% 
  summarise(
    n = n(),
    mean_duration = mean(duration),
    sd_duration = sd(duration),
    median_duration = duration(median(dseconds(duration))),
    median_duration_numeric = as.numeric(median(duration)),
    .groups = "drop"
  )

duration_summary_per_polishing <- 
  duration_summary_per_process %>% 
  group_by(polishing, cluster_per_polishing_file, mode, max_reads_per_cluster) %>% 
  summarise(
    n_processes = n(),
    total_median_duration = sum(median_duration),
    total_mean_duration = sum(mean_duration),
    .groups = "drop"
  )

```

## Total execution time pipeline

```{r}
pipeline_start_per_condition <-
  trace_files  %>% 
  filter(max_reads_per_cluster <= 50) %>% 
  filter(cluster_per_polishing_file <= 50) %>% 
  group_by(polishing, mode, max_reads_per_cluster, cluster_per_polishing_file) %>% 
  arrange(submit) %>% 
  slice_head(n = 1) %>% 
  select(
    start = submit, 
    polishing, 
    mode, 
    cluster_per_polishing_file, 
    max_reads_per_cluster, 
    process_order
  )

pipeline_end_per_condition <-
  trace_files  %>% 
  filter(max_reads_per_cluster <= 50) %>% 
  filter(cluster_per_polishing_file <= 50) %>% 
  group_by(polishing, mode, max_reads_per_cluster, cluster_per_polishing_file) %>% 
  arrange(desc(submit)) %>% 
  slice_head(n = 1) %>% 
  select(
    end = submit, 
    polishing, 
    mode, 
    cluster_per_polishing_file, 
    max_reads_per_cluster, 
    process_order
  )

pipeline_duration <- 
  pipeline_start_per_condition %>% 
  inner_join(pipeline_end_per_condition) %>% 
  mutate(
    duration = end - start
  )


write_tsv(pipeline_duration, file = "../benchmarking/out/pipeline_duration.tsv")
```

## Execution reduction per condition - GPU reference based

```{r}
reference_gpu <- duration_summary_per_polishing %>%
  filter(polishing == "Reference-based" & mode == "gpu") %>%
  select(
    cluster_per_polishing_file, 
    max_reads_per_cluster, 
    reference_mean = total_mean_duration, 
    reference_median = total_median_duration
  )

# Step 3: Join and compute x-fold reduction
duration_comparison_gpu <- duration_summary_per_polishing %>%
  left_join(reference_gpu, by = c("cluster_per_polishing_file", "max_reads_per_cluster")) %>%
  mutate(
    x_fold_reduction_median = total_median_duration / reference_median,
    x_fold_reduction_mean = total_mean_duration / reference_mean,
  )
```

### Table

```{r}
table_data_duration_comparison_gpu <- duration_comparison_gpu %>%
  mutate(
    total_mean_duration = round(total_mean_duration, 1),
    reference_mean = round(reference_mean, 1),
    x_fold_reduction_mean = round(x_fold_reduction_mean, 2)
  ) %>%
  select(
    Polishing = polishing,
    Mode = mode,
    `Clusters/File` = cluster_per_polishing_file,
    `Max Reads/Cluster` = max_reads_per_cluster,
    `Mean Duration (s)` = total_mean_duration,
    `GPU Reference (s)` = reference_mean,
    `x-Fold Slower` = x_fold_reduction_mean
  )

publication_table_duration_comparison_gpu <- flextable(table_data_duration_comparison_gpu) %>%
  theme_booktabs() %>%
  autofit() %>%
  set_caption("Mean durations and x-fold slowdown compared to GPU Reference-based polishing.") %>%
  align(align = "center", part = "all") %>%
  bold(part = "header") %>%
  fontsize(size = 11, part = "all") %>%
  padding(padding = 5, part = "all") %>% 
  theme_zebra() %>%
  border_remove() %>%
  border_outer(part = "all", border = fp_border(color = "black", width = 1)) %>%  
  border_inner_v(border = fp_border(color = "grey60", width = 0.5))%>%
  border_inner_h(border = fp_border(color = "grey60", width = 0.5)) 


doc_duration_comparison_gpu <- read_docx() %>%
  body_add_flextable(publication_table_duration_comparison_gpu) %>%
  body_add_par("*Note: x-Fold Slower is relative to GPU runtime at same configuration.", style = "Normal")

print(doc_duration_comparison_gpu, target = "../benchmarking/out/SuppTable_duration_comparison_gpu.docx")

```

## Execution reduction per condition - CPU reference based

```{r}
reference_cpu <- duration_summary_per_polishing %>%
  filter(polishing == "Reference-based" & mode == "cpu") %>%
  select(
    cluster_per_polishing_file, 
    max_reads_per_cluster, 
    reference_mean = total_mean_duration, 
    reference_median = total_median_duration
  )

# Step 3: Join and compute x-fold reduction
duration_comparison_cpu <- duration_summary_per_polishing %>%
  filter(mode == "cpu") %>% 
  left_join(reference_cpu, by = c("cluster_per_polishing_file", "max_reads_per_cluster")) %>%
  mutate(
    x_fold_reduction_median = total_median_duration / reference_median,
    x_fold_reduction_mean = total_mean_duration / reference_mean,
  )
```

### Table

```{r}
table_data_duration_comparison_cpu <- duration_comparison_cpu %>%
  mutate(
    total_mean_duration = round(total_mean_duration, 1),
    reference_mean = round(reference_mean, 1),
    x_fold_reduction_mean = round(x_fold_reduction_mean, 2)
  ) %>%
  select(
    Polishing = polishing,
    Mode = mode,
    `Clusters/File` = cluster_per_polishing_file,
    `Max Reads/Cluster` = max_reads_per_cluster,
    `Mean Duration (s)` = total_mean_duration,
    `CPU Reference (s)` = reference_mean,
    `x-Fold Slower` = x_fold_reduction_mean
  )

publication_table_duration_comparison_cpu <- flextable(table_data_duration_comparison_cpu) %>%
  theme_booktabs() %>%
  autofit() %>%
  set_caption("Mean durations and x-fold slowdown compared to CPU Reference-based polishing.") %>%
  align(align = "center", part = "all") %>%
  bold(part = "header") %>%
  fontsize(size = 11, part = "all") %>%
  padding(padding = 5, part = "all") %>% 
  theme_zebra() %>%
  border_remove() %>%
  border_outer(part = "all", border = fp_border(color = "black", width = 1)) %>%  
  border_inner_v(border = fp_border(color = "grey60", width = 0.5))%>%
  border_inner_h(border = fp_border(color = "grey60", width = 0.5)) 

doc_duration_comparison_cpu <- read_docx() %>%
  body_add_flextable(publication_table_duration_comparison_cpu) %>%
  body_add_par("*Note: x-Fold Slower is relative to CPU runtime at same configuration.", style = "Normal")

print(doc_duration_comparison_cpu, target = "../benchmarking/out/SuppTable_duration_comparison_cpu.docx")

```


## variant calling - mutserve

```{r}
mutserve_files_filtered <- 
  mutserve_files %>% 
  filter(max_reads_per_cluster <= 50) %>% 
  filter(cluster_per_polishing_file <= 50)

mutserve_variant_calling_summary <- 
  mutserve_files_filtered %>% 
  group_by(polishing, mode, max_reads_per_cluster, cluster_per_polishing_file) %>% 
  summarise(
    n_variants = n(),
    mean_variant_level = mean(`Variant-level`),
    sd_variant_level = sd(`Variant-level`)
  )

```

### Table

```{r}
table_data_mutserve_variant_calling_summary <- mutserve_variant_calling_summary %>%
  mutate(
    mean_variant_level = round(mean_variant_level * 100, 3),
    sd_variant_level = round(sd_variant_level * 100, 3),
  ) %>%
  select(
    Polishing = polishing,
    Mode = mode,
    `Clusters/File` = cluster_per_polishing_file,
    `Max Reads/Cluster` = max_reads_per_cluster,
    `Number of variants [n]` = n_variants,
    `Mean Variant Level (%)` = mean_variant_level,
    `SD Variant Level (%)` = sd_variant_level
  )

publication_table_mutserve_variant_calling_summary <- flextable(table_data_mutserve_variant_calling_summary) %>%
  theme_booktabs() %>%
  autofit() %>%
  fit_to_width(max_width = 6.5) %>%
  set_caption("Variant-level error metrics across polishing strategies.") %>%
  align(align = "center", part = "all") %>%
  bold(part = "header") %>%
  fontsize(size = 11, part = "all") %>%
  padding(padding = 5, part = "all") %>% 
  theme_zebra() %>%
  border_remove() %>%
  border_outer(part = "all", border = fp_border(color = "black", width = 1)) %>%  
  border_inner_v(border = fp_border(color = "grey60", width = 0.5))%>%
  border_inner_h(border = fp_border(color = "grey60", width = 0.5)) 

doc_mutserve_variant_calling_summary <- read_docx() %>%
  body_add_flextable(publication_table_mutserve_variant_calling_summary)

print(doc_mutserve_variant_calling_summary, target = "../benchmarking/out/SuppTable_mutserve_variant_calling_summary.docx")

```

## RAM consumption 
```{r}
# Create a clean label mapping
condition_labels <- c(
  "Reference-based_CPU" = "Reference-based CPU",
  "Reference-based_GPU" = "Reference-based GPU",
  "POA-based_GPU" = "POA-based GPU",
  "POA-based_CPU" = "POA-based CPU"
)

# Add a combined condition column
ram_summary <- trace_files_parsed %>%
  mutate(
    cluster_per_polishing_file = as.factor(cluster_per_polishing_file),
    max_reads_per_cluster = as.factor(max_reads_per_cluster),
    condition = paste(polishing, mode, sep = "_")
  ) %>%
  group_by(condition, cluster_per_polishing_file, max_reads_per_cluster, polishing, mode) %>%
  summarise(
    mean_ram_mb = mean(peak_rss_mb), 
    n = n(),
    .groups = "drop"
  ) %>% 
  mutate(
    n = ifelse(str_detect(polishing, "Reference"), n /4 , n),
  )

# Compute baseline per cluster size (GPU + Reference-based)
baseline_per_cluster <- ram_summary %>%
  filter(condition == "Reference-based_CPU") %>%
  select(cluster_per_polishing_file, max_reads_per_cluster, baseline_ram = mean_ram_mb)

# Join and compute relative RAM
ram_summary <- ram_summary %>%
  left_join(baseline_per_cluster, by = c("cluster_per_polishing_file", "max_reads_per_cluster")) %>%
  mutate(
    ram_relative_to_baseline = mean_ram_mb / baseline_ram
  )

# Apply factor with nice labels
ram_summary <- ram_summary %>%
  mutate(
    condition = factor(
      condition, 
      levels = c("Reference-based_CPU", "Reference-based_GPU", "POA-based_GPU", "POA-based_CPU"),
      labels = condition_labels
    ),
    max_reads_per_cluster = factor(max_reads_per_cluster)  # optional: keep numeric
  )

```

### Table
```{r}

# Prepare table data
table_data_ram_comparison <- ram_summary %>%
  mutate(
    mean_ram_mb = round(mean_ram_mb, 1),
    ram_relative_to_baseline = round(ram_relative_to_baseline, 2),
  ) %>%
  select(
    Polishing = polishing,
    Mode = mode,
    `Clusters/File` = cluster_per_polishing_file,
    `Max Reads/Cluster` = max_reads_per_cluster,
    `Mean RAM (MB)` = mean_ram_mb,
    `Relative RAM*` = ram_relative_to_baseline,
  )

# Create publication-ready flextable
publication_table_ram_comparison <- flextable(table_data_ram_comparison) %>%
  theme_booktabs() %>%
  autofit() %>%
  set_caption("Absolute and relative RAM consumption per polishing step across conditions.") %>%
  align(align = "center", part = "all") %>%
  bold(part = "header") %>%
  fontsize(size = 11, part = "all") %>%
  padding(padding = 5, part = "all") %>% 
  theme_zebra() %>%
  border_remove() %>%
  border_outer(part = "all", border = fp_border(color = "black", width = 1)) %>%  
  border_inner_v(border = fp_border(color = "grey60", width = 0.5)) %>%
  border_inner_h(border = fp_border(color = "grey60", width = 0.5))

# Export to Word
doc_ram_comparison <- read_docx() %>%
  body_add_flextable(publication_table_ram_comparison) %>%
  body_add_par("*Note: Relative RAM is normalized to Reference-based CPU for the same configuration.", style = "Normal")

print(doc_ram_comparison, target = "../benchmarking/out/SuppTable_ram_comparison.docx")

```


# plots

## Total execution time
### Create panels
```{r}
make_panel_pipeline_duration <- function(polishing_filter, mode_filter){
  # Filter data
  counts <- 
    ram_summary %>% 
    filter(polishing == polishing_filter, mode == mode_filter) %>% 
    filter( max_reads_per_cluster == 30)
  
  df <- pipeline_duration %>% 
    filter(polishing == polishing_filter, mode == mode_filter) %>% 
    mutate(
      cluster_per_polishing_file = as.factor(cluster_per_polishing_file),
      max_reads_per_cluster = as.factor(max_reads_per_cluster)
    )
  
  ggplot(df, aes(
    x = cluster_per_polishing_file,
    y = duration,
    fill = max_reads_per_cluster
  )) +
    geom_bar(
      stat = "identity", 
      position = "dodge"
    ) +
    facet_grid(
      cols = vars(polishing)
    ) +
    # add text with number of processes per panel
    geom_text(
      data = counts,
      aes(
        y = max(df$duration) * 1.1,  # above bars
        label = paste("n =", n)
      ),
      size = 6
    ) +
    scale_y_continuous(
      expand = expansion(mult = c(0, 0.08))  # 20% extra space at the top
    ) +
    scale_fill_npg() +
    labs(
      x = "Cluster per polishing file",
      y = "Duration [s]", 
      fill = "Max reads per cluster"
    ) +
    theme_bw() +
    theme(
      axis.text = element_text(colour = "black"),
      text = element_text(size = 20, colour = "black"), 
      legend.position = "bottom"
    )
}


p.pipeline_duration_reference_cpu <- make_panel_pipeline_duration("Reference-based", "CPU")
p.pipeline_duration_reference_gpu <- make_panel_pipeline_duration("Reference-based", "GPU") 
p.pipeline_duration_poa_gpu <- make_panel_pipeline_duration("POA-based", "GPU") 
p.pipeline_duration_poa_cpu <- make_panel_pipeline_duration("POA-based", "CPU")

```

#### patchwork

```{r}
design_pipeline_duration <- "AA\nBB\nCC\nDD"

p.pipeline_duration_per_condition <- 
  wrap_plots(
    A = p.pipeline_duration_reference_cpu, 
    B = p.pipeline_duration_reference_gpu, 
    C = p.pipeline_duration_poa_gpu, 
    D = p.pipeline_duration_poa_cpu, 
    design = design_pipeline_duration,
    guides = "collect"
  ) +
  plot_annotation(tag_levels = 'A') +
  plot_layout(
    axes = "collect",
    axis_titles = "collect",
    guides = "collect"
  ) &
  theme(
    axis.title.y = element_text(margin = margin(l = 5, r = 5)),
    legend.position = "bottom"
  )

p.pipeline_duration_per_condition

ggsave(
  plot = p.pipeline_duration_per_condition,
  filename = "pipeline_duration_per_condition.jpeg", 
  path = "../benchmarking/out/",
  width = 12, 
  height = 12
)
```




## RAM consumption

### Create Panels
```{r}
make_panel_ram_consumption <- function(polishing_filter, mode_filter){
  trace_files_parsed %>% 
    filter(polishing == polishing_filter) %>% 
    filter(mode == mode_filter) %>% 
    mutate(
      cluster_per_polishing_file = as.factor(cluster_per_polishing_file),
      max_reads_per_cluster = as.factor(max_reads_per_cluster)
    ) %>%
    ggplot(aes(
      x = cluster_per_polishing_file,
      y = peak_rss_mb,
      fill = max_reads_per_cluster
    )) +
    geom_boxplot(
      position = position_dodge(width = 0.8),
      width = 0.7
    ) +
    facet_grid(
      cols = vars(reorder(process, process_order))
    ) +
    scale_fill_npg() +
    labs(
      x = "Cluster per polishing file",
      y = "RAM consumption [MB]", 
      fill = "Max reads per cluster"
    ) +
    theme_bw() +
    theme(
      axis.text = element_text(colour = "black"),
      text = element_text(size = 20, colour = "black"),
      legend.position = "bottom"
    )
}

p.ram_consumption_reference_cpu <- make_panel_ram_consumption("Reference-based", "CPU")
p.ram_consumption_reference_gpu <- make_panel_ram_consumption("Reference-based", "GPU") 
p.ram_consumption_poa_gpu <- make_panel_ram_consumption("POA-based", "GPU") 
p.ram_consumption_poa_cpu <- make_panel_ram_consumption("POA-based", "CPU")
```

#### patchwork
```{r}
design_ram_consumption <- "AA\nBB\nCD"

p.ram_consumption_per_condition <- 
  wrap_plots(
    A = p.ram_consumption_reference_cpu,
    B = p.ram_consumption_reference_gpu,
    C = p.ram_consumption_poa_gpu,
    D = p.ram_consumption_poa_cpu,
    design = design_ram_consumption,
    guides = "collect"
  ) +
  plot_annotation(tag_levels = 'A') +
  plot_layout(
    axes = "collect",
    axis_titles = "collect",
    guides = "collect"
  ) &
  theme(
    legend.position = "bottom"
  )

p.ram_consumption_per_condition

ggsave(
  plot = p.ram_consumption_per_condition,
  filename = "ram_consumption_per_condition.jpeg", 
  path = "../benchmarking/out/",
  width = 12, 
  height = 9
)

```

## RAM consumption - barplots
### Create Panels
```{r}
make_panel <- function(cluster_size) {
  annotation <- ram_summary %>% 
    filter(cluster_per_polishing_file == cluster_size) %>%
    slice_head(n = 1) %>% 
    pull(n)
  
  ram_summary %>% 
    filter(cluster_per_polishing_file == cluster_size) %>%
    ggplot(aes(
      x = condition,
      y = ram_relative_to_baseline,
      fill = max_reads_per_cluster
    )) +
    geom_bar(stat = "identity", position = position_dodge()) + 
    # geom_text() +
    annotate(
      geom = "text",
      label = paste("n =", annotation), 
      x = 1, y = 30, size = 6
    ) +
    coord_cartesian(ylim = c(0, 30)) +
    scale_y_continuous(breaks = seq(0, 30, 5)) +
    labs(
      x = "Condition",
      y = "RAM consumption relative to\nCPU Reference-based",
      fill = "Max reads per cluster"
    ) +
    theme_bw() +
    scale_fill_npg() +
    theme(
      text = element_text(size = 20, colour = "black"),
      axis.text = element_text(colour = "black"),
      axis.text.x = element_text(angle = 45, hjust = 1, colour = "black"),
      legend.position = "bottom"
    )
}


# Create three separate plots
p10 <- make_panel(10)
p20 <- make_panel(20)
p50 <- make_panel(50)

ram_comparison_labels_design <- "ABC"

# Combine them with patchwork
p.ram_comparison <- 
  wrap_plots(
    A = p10, 
    B = p20, 
    C = p50, 
    design = ram_comparison_labels_design, 
    guides = "collect"
  ) + 
  plot_annotation(
    tag_levels = "A",
  ) +
  plot_layout(
    axes = "collect",
    guides = "collect"
  ) & 
  theme(legend.position = "bottom")

p.ram_comparison

ggsave(
  plot = p.ram_comparison,
  filename = "ram_comparison.jpeg", 
  path = "../benchmarking/out/",
  width = 12, 
  height = 9
)

```

## polishing duration per condition and process

### Create panels
```{r}
make_panel_polishing_duration_per_condition <- function(polishing_filter, mode_filter){
  # Filter data
  counts <- 
    ram_summary %>% 
    filter(polishing == polishing_filter, mode == mode_filter) %>% 
    filter( max_reads_per_cluster == 30)
  
  df <- trace_files_parsed %>% 
    filter(polishing == polishing_filter, mode == mode_filter) %>% 
    mutate(
      cluster_per_polishing_file = as.factor(cluster_per_polishing_file),
      max_reads_per_cluster = as.factor(max_reads_per_cluster)
    )
  
  df %>% 
    filter(polishing == polishing_filter) %>% 
    filter(mode == mode_filter) %>% 
    mutate(
      cluster_per_polishing_file = as.factor(cluster_per_polishing_file),
      max_reads_per_cluster = as.factor(max_reads_per_cluster), 
    ) %>%
    ggplot(aes(
      x = cluster_per_polishing_file,
      y = duration,
      fill = max_reads_per_cluster
    )) +
    facet_grid(
      cols = vars(reorder(process, process_order))
    ) +
    geom_boxplot(
      position = position_dodge(width = 0.8),
      width = 0.7
    ) + 
    # add text with number of processes per panel
    geom_text(
      data = counts,
      aes(
        y = max(df$duration) * 1.1,  # above bars
        label = n
      ),
      size = 6
    ) +
    scale_y_continuous(
      expand = expansion(mult = c(0, 0.08))  # 20% extra space at the top
    ) +
    scale_fill_npg() +
    labs(
      x = "Cluster per polishing file",
      y = "Duration [s]", 
      fill = "Max reads per cluster"
    ) +
    theme_bw() +
    theme(
      axis.text = element_text(colour = "black"),
      text = element_text(size = 20, colour = "black")
    )
}

p.polishing_duration_per_condition_reference_cpu <- make_panel_polishing_duration_per_condition("Reference-based", "CPU")
p.polishing_duration_per_condition_reference_gpu <- make_panel_polishing_duration_per_condition("Reference-based", "GPU") 
p.polishing_duration_per_condition_poa_gpu <- make_panel_polishing_duration_per_condition("POA-based", "GPU") 
p.polishing_duration_per_condition_poa_cpu <- make_panel_polishing_duration_per_condition("POA-based", "CPU")


```

#### patchwork

```{r}
design_polishing_duration_per_condition <- "A\nB\nC\nD"

p.polishing_duration_per_condition <- 
  wrap_plots(
    A = p.polishing_duration_per_condition_poa_cpu, 
    B = p.polishing_duration_per_condition_reference_cpu, 
    C = p.polishing_duration_per_condition_poa_gpu, 
    D = p.polishing_duration_per_condition_reference_gpu, 
    design = design_polishing_duration_per_condition,
    # heights = c(1, 1),
    # widths = c(3, 11),
    guides = "collect"
  ) +
  plot_annotation(
    tag_levels = 'A', 
  ) +
  plot_layout(
    axes = "collect",
    axis_titles = "collect",
    guides = "collect"
  ) &
  theme(
    axis.title.y = element_text(margin = margin(l = 5, r = 5)),
    legend.position = "bottom",
    # strip.text = element_text(size = 14)
  )

p.polishing_duration_per_condition

ggsave(
  plot = p.polishing_duration_per_condition,
  filename = "polishing_duration_per_condition.jpeg", 
  path = "../benchmarking/out/",
  dpi = 600,
  width = 12, 
  height = 12
)
```

## Variant detection

```{r}
p.variant_detection_changes <-
  mutserve_files %>% 
  ggplot(aes(Position, `Variant-level`, color = mode)) +
  geom_point() +
  facet_wrap(vars(polishing, mode)) +
  theme_bw() +
  scale_color_manual(values = c("cpu" = "#E69F00", "gpu" = "#56B4E9")) +
  theme(
    text = element_text(size = 22, colour = "black"),
    axis.text = element_text(colour = "black"),
  )

ggsave(
  plot = p.variant_detection_changes,
  filename = "variant_detection_changes.jpeg", 
  path = "../benchmarking/out/",
  width = 12, 
  height = 9
)


```

# Lunch seminar 
## RAM
```{r}
cluster_per_file <- 20
reads_per_cluster <- 30
annotation <- ram_summary %>% 
  filter(cluster_per_polishing_file == cluster_per_file) %>%
  slice_head(n = 1) %>% 
  pull(n)

ram_summary %>% 
  filter(cluster_per_polishing_file == cluster_per_file) %>%
  filter(max_reads_per_cluster == max_reads_per_cluster) %>%
  ggplot(aes(
    x = condition,
    y = ram_relative_to_baseline, 
    fill = condition
  )) +
  geom_bar(stat = "identity", position = position_dodge(), show.legend = FALSE) + 
  annotate(
    geom = "text",
    label = paste("n =", annotation), 
    x = 0.8, y = 25, size = 8
  ) +
  coord_cartesian(ylim = c(0, 25)) +
  scale_y_continuous(
    breaks = seq(0, 30, 5),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    x = "Condition",
    y = "RAM consumption relative to\nCPU Reference-based",
    fill = "Max reads per cluster"
  ) +
  theme_bw() +
  scale_fill_npg() +
  theme(
    text = element_text(size = 28, colour = "black"),
    axis.text = element_text(colour = "black"),
    axis.text.x = element_text(angle = 30, hjust = 1, colour = "black"),
    legend.position = "bottom"
  )

ggsave(
  filename = "lunch_seminar_RAM.jpeg", 
  path = "../lunch_seminar/out/",
  dpi = 600,
  width = 12, 
  height = 9
)

```

## Pipeline duration 
```{r}
pipeline_duration %>% 
  filter(cluster_per_polishing_file == cluster_per_file) %>%
  filter(reads_per_cluster == max_reads_per_cluster) %>%
  mutate(
    cluster_per_polishing_file = as.factor(cluster_per_polishing_file),
    max_reads_per_cluster = as.factor(max_reads_per_cluster)
  ) %>% 
  ggplot(aes(
    x = reorder(paste(polishing, mode), duration),
    y = duration,
    fill = reorder(paste(polishing, mode), duration)
  )) +
  geom_bar(
    stat = "identity", 
    position = "dodge", 
    show.legend = FALSE
  ) +
  annotate(
    geom = "text",
    label = paste("n =", annotation), 
    x = 0.7, y = 4000, size = 8
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.08))  # 20% extra space at the top
  ) +
  scale_fill_npg() +
  labs(
    x = "Condition",
    y = "Duration [s]", 
    fill = "Max reads per cluster"
  ) +
  theme_bw() +
  scale_fill_npg() +
  theme(
    text = element_text(size = 28, colour = "black"),
    axis.text = element_text(colour = "black"),
    axis.text.x = element_text(angle = 30, hjust = 1, colour = "black"),
    legend.position = "bottom"
  )

ggsave(
  filename = "lunch_seminar_duration.jpeg", 
  path = "../lunch_seminar/out/",
  dpi = 600,
  width = 12, 
  height = 9
)
```

