---
title: "early_stopping_QC"
format: html
editor: visual
---

# library

```{r}
library(tidyverse)
library(flextable)
library(officer)
library(patchwork)
```

# funtions

```{r}
read_cluster_stats <- function(path){
  # extract timestamt "20260202_1735"
  timestamp <- str_extract(path, "\\d{8}_\\d{4}")
  
  read_tsv(path) %>%
    mutate(
      timestamp = as.POSIXct(timestamp, format = "%Y%m%d_%H%M", tz = "UTC"),
    )
}

read_filtered_read_stats <- function(path){
  barcode <- str_extract(path, "barcode\\d+")
  timestamp <- str_extract(path, "\\d{8}_\\d{4}")
  
  
  read_tsv(path) %>%
    filter(format == "count") %>% 
    select(reads_found, reads_on_target, reads_filtered) %>%
    pivot_longer(
      cols = c(reads_found, reads_on_target, reads_filtered), 
      names_to = "filter_step",
      values_to = "n_reads"
    ) %>% 
    mutate(
      timestamp = as.POSIXct(timestamp, format = "%Y%m%d_%H%M", tz = "UTC"),
      barcode = barcode
    )
  
}

read_extracted_umi_stats <- function(path){
  barcode <- str_extract(path, "barcode\\d+")
  timestamp <- str_extract(path, "\\d{8}_\\d{4}")
  
  read_tsv(path) %>%
    select(detected_forward_strands, detected_reverse_strands, total_reads, included_reads) %>% 
    pivot_longer(
      cols = c(detected_forward_strands, detected_reverse_strands, total_reads, included_reads), 
      names_to = "umi_metric",
      values_to = "n_counts"
    ) %>%
    mutate(
      timestamp = as.POSIXct(timestamp, format = "%Y%m%d_%H%M", tz = "UTC"),
      barcode = barcode
    )
}

read_variant_calling <- function(path){
  barcode <- str_extract(path, "barcode\\d+")
  mode <- str_remove(str_extract(path, "full\\/|early\\/"), "/")
  
  read_tsv(path) %>%
    mutate(
      barcode = barcode, 
      mode = mode
    ) %>% 
    select(
      barcode, 
      position = Position,
      reference = Reference,
      variant = Variant,
      variant_level = `Variant-level`
    )
}

read_summary <- function(path){
  barcode <- str_extract(path, "barcode\\d+")
  mode <- str_remove(str_extract(path, "full\\/|early\\/"), "/")  
  read_tsv(path) %>%
    mutate(
      mode = mode,
      barcode = barcode
    )
}
```

# paths

```{r}
cluster_stats_paths <- list.files(
  path =  "early_stopping/data/early/cluster_stats", 
  pattern = "summary_cluster_stats.tsv", 
  recursive = TRUE, 
  full.names = TRUE
)

filtered_read_stats_paths <- list.files(
  path =  "early_stopping/data/early/", 
  pattern = "filter_reads_stats.tsv", 
  recursive = TRUE, 
  full.names = TRUE
)

extracted_umis_stats_paths <- list.files(
  path =  "early_stopping/data/early", 
  pattern = "extracted_umis_barcode", 
  recursive = TRUE, 
  full.names = TRUE
)

variant_calling_pahts <- list.files(
  path =  "early_stopping/data/", 
  pattern = "consensus_parsed.txt$", 
  recursive = TRUE, 
  full.names = TRUE
)

cluster_stats_full_path <- "early_stopping/data/full/cluster_stats/summary_cluster_stats.tsv"
sample_sheet_path <- "early_stopping/data/sample_sheet/sample_sheet.tsv"
```

# load data

```{r, message=FALSE}
cluster_stats <- lapply(cluster_stats_paths, read_cluster_stats ) %>%
  bind_rows()


filtered_read_stats <- lapply(filtered_read_stats_paths, read_filtered_read_stats ) %>%
  bind_rows()

extracted_umis_stats <- lapply(extracted_umis_stats_paths, read_extracted_umi_stats ) %>%
  bind_rows()

variant_calling_stats_early_paths <- variant_calling_pahts[ str_detect(variant_calling_pahts, "early/") ]
variant_calling_stats_full_paths <- variant_calling_pahts[ str_detect(variant_calling_pahts, "full/") ]

variant_calling_stats_early <- lapply(variant_calling_stats_early_paths, read_variant_calling ) %>%
  bind_rows()

variant_calling_stats_full <- lapply(variant_calling_stats_full_paths, read_variant_calling ) %>%
  bind_rows()

cluster_stats_full <- read_tsv(cluster_stats_full_path) %>% 
  mutate(
    mode = "full"
  )
sample_sheet <- read_tsv(sample_sheet_path)
```

# parse tables

## start time

```{r}
start_time <- min(filtered_read_stats$timestamp)
```

## extracted umis parsed

```{r}
# accumulate reads over time
extracted_umis_stats_accumulated <- 
  extracted_umis_stats %>%
  group_by(barcode, umi_metric) %>%
  arrange(timestamp) %>%
  mutate(
    accumulated_counts = cumsum(n_counts), 
    runtime_hours = as.numeric(difftime(timestamp, start_time, units = "hours")),
  ) %>%
  inner_join(sample_sheet) %>%
  ungroup()
```

## filtered read stats parsed

```{r}
filtered_read_stats_accumulated <- 
  filtered_read_stats %>%
  group_by(barcode, filter_step) %>%
  arrange(timestamp) %>%
  mutate(
    accumulated_reads = cumsum(n_reads),
    runtime_hours = as.numeric(difftime(timestamp, start_time, units = "hours")),
  ) %>%
  inner_join(sample_sheet) %>% 
  ungroup()
```

## paste timestamp to date time

```{r}
cluster_stats_parsed <- 
  cluster_stats %>%
  mutate(
    runtime_hours = as.numeric(difftime(timestamp, start_time, units = "hours"))
  ) %>% 
  rename(
    barcode = sample
  ) %>%
  inner_join(sample_sheet)
```

### save for main figure

```{r}
cluster_stats_parsed %>%
  write_tsv("main_figure/data/early_stopping_cluster_stats_parsed.tsv")
```

## variant calling performance

calculate the correlation and performance metrics of variant calling between early stopping and full run \### join data

```{r}
variant_calling_comparison <-
  variant_calling_stats_early %>%
  full_join(
    variant_calling_stats_full,
    by = c("barcode", "position", "reference", "variant"),
    suffix = c("_early", "_full")
  ) %>% 
  inner_join(sample_sheet) %>%
  mutate(
    variant_level_early = ifelse(is.na(variant_level_early), 0, variant_level_early),
    variant_level_full = ifelse(is.na(variant_level_full), 0, variant_level_full)
  )
```

#### save for main figure

```{r}
variant_calling_comparison %>%
  write_tsv("main_figure/data/early_stopping_variant_calling_comparison.tsv")
```

### correlation

```{r}
variant_calling_correlation <- 
  variant_calling_comparison %>%
  group_by(sample) %>%
  summarise(
    correlation = cor(variant_level_early, variant_level_full, method = "pearson"),
    # calculate R-squared coeefficient of determination using a regression model
    r_squared = summary(lm(variant_level_early ~ variant_level_full))$r.squared,
    # rmse = sqrt(mean((variant_level_early - variant_level_full)^2)),
    # mae = mean(abs(variant_level_early - variant_level_full))
  ) %>%
  ungroup()

# correlation summary
variant_calling_correlation_summary <- 
  variant_calling_correlation %>%
  summarise(
    mean_correlation = mean(correlation),
    median_correlation = median(correlation),
    sd_correlation = sd(correlation)
  )

```

### Number of variants

```{r}
variant_calling_variant_counts <- 
  variant_calling_comparison %>%
  group_by(sample) %>%
  summarise(
    n_variants_early = sum(variant_level_early > 0),
    n_variants_full = sum(variant_level_full > 0)
  ) %>%
  ungroup()
```

### performance metrics

```{r}
length <- 5104
variant_calling_performance <- 
  variant_calling_comparison %>%
  mutate(
    true_positive = variant_level_early > 0 & variant_level_full > 0,
    false_positive = variant_level_early > 0 & variant_level_full == 0,
    false_negative = variant_level_early == 0 & variant_level_full > 0,
  ) %>%
  group_by(sample) %>%
  summarise(
    true_positive = sum(true_positive),
    false_positive = sum(false_positive),
    false_negative = sum(false_negative),
    true_negative = length - (true_positive + false_positive + false_negative),  # assuming total possible variants is 5104
    sensitivity = true_positive / (true_positive + false_negative),
    specificity = true_negative / (true_negative + false_positive),
    precision = true_positive / (true_positive + false_positive),
    # accuracy = (true_positive + true_negative) / length
  ) %>%
  ungroup()

variant_calling_performance
```

### combine

```{r}
variant_calling_summary <- 
  variant_calling_correlation %>%
  inner_join(variant_calling_performance)
```

### write table

```{r}
ft_variant_calling_summary <- 
  variant_calling_summary %>% 
  mutate(
    correlation = round(correlation, 3),
    r_squared = round(r_squared, 3),
    sensitivity = round(sensitivity, 3),
    specificity = round(specificity, 3),
    precision = round(precision, 3),
  ) %>% 
  flextable() %>%
  set_header_labels(
    correlation = "Pearson's r",
    r_squared = "R-squared",
    true_positive = "True Positives",
    false_positive = "False Positives",
    false_negative = "False Negatives",
    true_negative = "True Negatives",
    sensitivity = "Sensitivity",
    specificity = "Specificity",
    precision = "Precision"
  ) %>%
  align(align = "center", part = "all") %>%
  bold(part = "header") %>%
  fontsize(size = 11, part = "all") %>%
  padding(padding = 5, part = "all") %>% 
  theme_zebra() %>%
  border_remove() %>%
  border_outer(part = "all", border = fp_border(color = "black", width = 1)) %>%  
  border_inner_v(border = fp_border(color = "grey60", width = 0.5))%>%
  border_inner_h(border = fp_border(color = "grey60", width = 0.5))

# Export to Word
doc_variant_calling_summary <- read_docx() %>%
  body_add_par("Variant Calling Performance Summary", style = "heading 1") %>%
  body_add_flextable(ft_variant_calling_summary) %>%
  body_add_par("", style = "Normal")  # optional spacer

print(doc_variant_calling_summary, target = "early_stopping/out/Variant_Calling_Performance_Summary.docx")
```

## number of cluster per sample in both conditions

### latest clustering early stopping

```{r}
cluster_stats_latest_early <- 
  cluster_stats_parsed %>%
  group_by(sample) %>%
  arrange(timestamp) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  select(sample, n_cluster, runtime_hours)

cluster_stats_latest_early_summary <- 
  cluster_stats_latest_early %>%
  summarise(
    mean_clusters_early = mean(n_cluster),
    median_clusters_early = median(n_cluster),
    sd_clusters_early = sd(n_cluster),
    total_clusters_early = sum(n_cluster)
  )
```

### compare to full run

```{r}
# manually set runtime hours to 29h (first run 19h for cluster monitoring and second run 10h for live clustering)
cluster_stats_full_parsed <- 
  cluster_stats_full %>%
  select(
    barcode = sample, 
    n_cluster_full= n_cluster
  ) %>%
  mutate(
    runtime_hours_full = 19.3
  ) %>% 
  inner_join(sample_sheet)

# full cluster summary
cluster_stats_full_summary <- 
  cluster_stats_full_parsed %>%
  summarise(
    mean_clusters_full = mean(n_cluster_full),
    median_clusters_full = median(n_cluster_full),
    sd_clusters_full = sd(n_cluster_full),
    total_clusters_full = sum(n_cluster_full)
  )

cluster_stats_comparison <-
cluster_stats_latest_early %>%
  left_join(cluster_stats_full_parsed, by = "sample") %>%
  mutate(
    n_cluster_full = ifelse(is.na(n_cluster_full), 0, n_cluster_full), 
    relative_cluster_number = n_cluster / n_cluster_full,
    relative_cluster_number_percent = round(relative_cluster_number * 100, 0),
    relative_runtime_hours = runtime_hours / runtime_hours_full,
    relative_runtime_hours_percent = round(relative_runtime_hours * 100, 0)
  )
```

#### write table

```{r}
ft_cluster_stats_comparison <- 
  cluster_stats_comparison %>% 
  select(sample, runtime_hours_full, n_cluster_full, runtime_hours, n_cluster, relative_cluster_number_percent, relative_runtime_hours_percent) %>% 
  mutate(
    runtime_hours = round(runtime_hours, 2),
  ) %>% 
  rename(
    Sample = sample,
    `Clusters (Early Stopping)` = n_cluster,
    `Clusters (Full Run)` = n_cluster_full,
    `Relative Clusters (%)` = relative_cluster_number_percent,
    `Runtime (Early Stopping, hours)` = runtime_hours,
    `Runtime (Full Run, hours)` = runtime_hours_full,
    `Relative Runtime (%)` = relative_runtime_hours_percent
  ) %>%
  flextable() %>%
  align(align = "center", part = "all") %>%
  bold(part = "header") %>%
  fontsize(size = 11, part = "all") %>%
  padding(padding = 5, part = "all") %>% 
  theme_zebra() %>%
  border_remove() %>%
  border_outer(part = "all", border = fp_border(color = "black", width = 1)) %>%  
  border_inner_v(border = fp_border(color = "grey60", width = 0.5))%>%
  border_inner_h(border = fp_border(color = "grey60", width = 0.5))

# Export to Word
doc_cluster_stats_comparison <- read_docx() %>%
  body_add_par("Cluster Statistics Comparison between Early Stopping and Full Run", style = "heading 1") %>%
  body_add_flextable(ft_cluster_stats_comparison) %>%
  body_add_par("", style = "Normal")  # optional spacer

print(doc_cluster_stats_comparison, target = "early_stopping/out/Cluster_Stats_Comparison.docx")
```

# plots

## variant level correlation

Take the faceted figure with correlations explaining the sample with less cluster has more residual noise

```{r}
p.variant_level_plot <-
  variant_calling_comparison %>%
  ggplot(aes(x = variant_level_full, y = variant_level_early)) +
  facet_wrap(~sample, ncol = 2) +extracted_umis_stats
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey80") +
  geom_text(
    data = variant_calling_correlation,
    aes(
      x = 0.1, 
      y = 0.98, 
      # paste R-saquared value with 4 decimal places
      label = paste0("R^2 == ", round(r_squared, 3))
    ),
    parse = TRUE,
    size = 5
  ) +
  # add number of variants per condition
  geom_text(
    data = variant_calling_variant_counts,
    aes(
      x = 0.17, 
      y = 0.85, 
      label = paste0("Variants = ", n_variants_early, " [n]")
    ),
    size = 5
  ) +
  # make percentage scale
  scale_x_continuous(labels = scales::percent, breaks = seq(0, 1, 0.2)) +
  scale_y_continuous(labels = scales::percent, breaks = seq(0, 1, 0.2)) +
  labs(
    # title = "Variant Level Correlation between Early Stopping and Full Run",
    x = "Variant Level (Full Run)",
    y = "Variant Level (Early Stopping)"
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 18),
  )

p.variant_level_plot

ggsave( 
  plot = p.variant_level_plot, 
  filename = "variant_level_plot.jpg", 
  path = "early_stopping/out", 
  width = 12, 
  height = 10, 
  dpi = 300
)
```

## accumulates reads, extracted umis and cluster over time

```{r}
p.accumulated_reads <-
  filtered_read_stats_accumulated %>%
  filter(filter_step == "reads_found") %>% 
  ggplot(aes(x = runtime_hours, y = accumulated_reads, color = sample)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    # title = "Accumulated Reads over Time",
    x = "Runtime (hours)",
    y = "Reads [n]",
    color = "Sample"
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 16),
    legend.position = "bottom"
  )

p.accumulated_reads

ggsave( 
  plot = p.accumulated_reads, 
  filename = "accumulated_reads_over_time.jpg", 
  path = "early_stopping/out", 
  width = 12, 
  height = 9, 
  dpi = 300
)
```

## accumulated extracted umis

```{r}
p.accumulated_umis <-
  extracted_umis_stats_accumulated %>%
  filter(umi_metric == "included_reads") %>% 
  ggplot(aes(x = runtime_hours, y = accumulated_counts, color = sample)) +
  geom_line() +
  geom_point() +
  labs(
    # title = "Accumulated Extracted UMIs over Time",
    x = "Runtime (hours)",
    y = "Extracted UMIs [n]",
    color = "Sample"
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 16),
    legend.position = "bottom"
  )

p.accumulated_umis

ggsave( 
  plot = p.accumulated_umis, 
  filename = "accumulated_extracted_umis_over_time.jpg", 
  path = "early_stopping/out", 
  width = 12, 
  height = 9, 
  dpi = 300
)
```

## accumulated clusters

```{r}
p.accumulated_clusters <-
  cluster_stats_parsed %>%
  ggplot(aes(x = runtime_hours, y = n_cluster, color = sample)) +
  geom_line(alpha = 0.3) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  labs(
    # title = "Accumulated Clusters over Time",
    x = "Runtime (hours)",
    y = "Clusters [n]",
    color = "Sample"
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 20), 
    legend.position = "bottom"
  )

p.accumulated_clusters

ggsave( 
  plot = p.accumulated_clusters, 
  filename = "accumulated_clusters_over_time.jpg", 
  path = "early_stopping/out", 
  width = 12, 
  height = 9, 
  dpi = 300
)
```

### combined plot

```{r}
design <- "A\nB\nC"

pw.combined_accumulated <- 
  wrap_plots(
    A = p.accumulated_reads,
    B = p.accumulated_umis,
    C = p.accumulated_clusters,
    design = design
  ) +
  plot_layout(
    axis_titles = "collect", 
    guides = "collect"
  ) + 
  plot_annotation(tag_levels = "A")

pw.combined_accumulated

ggsave( 
  plot = pw.combined_accumulated, 
  filename = "combined_accumulated_plots_over_time.jpg", 
  path = "early_stopping/out", 
  width = 12, 
  height = 18, 
  dpi = 300
)
```

### combined plot without umi

```{r}
design_wo_umi <- "A\nB"

pw.combined_accumulated_wo_umi <- 
  wrap_plots(
    A = p.accumulated_reads,
    B = p.accumulated_clusters,
    design = design_wo_umi
  ) +
  plot_layout(
    axis_titles = "collect", 
    guides = "collect", 
  ) + 
  plot_annotation(tag_levels = "A")

pw.combined_accumulated_wo_umi

ggsave( 
  plot = pw.combined_accumulated_wo_umi, 
  filename = "combined_accumulated_plots_over_time_wo_umi.jpg", 
  path = "early_stopping/out", 
  width = 12, 
  height = 18, 
  dpi = 300
)
```
