---
title: "Karst_benchmarking"
format: html
editor: visual
---

# library

```{r}
library(tidyverse)
library(flextable)
library(officer)
```

# file paths

```{r}
abundance_table_umi_poa_path <-  "Karst/context/poa/abundance_table_genus.tsv"
abundance_table_umi_reference_path <-  "Karst/context/reference_based/abundance_table_genus.tsv"
abundance_table_karst_path <-  "Karst/supplementary material/relative_abundance_zymo.txt"
abundance_table_karst_new_path <-  "Karst/karst/abundance_table_genus.tsv"
```

# read files

```{r}
abundance_table_umi_poa <- read_tsv(abundance_table_umi_poa_path) %>% 
  mutate(
    pipeline = "umi-pipeline-nf", 
    polishing = "POA-based"
  )

abundance_table_umi_reference <- read_tsv(abundance_table_umi_reference_path) %>% 
  mutate(
    pipeline = "umi-pipeline-nf", 
    polishing = "Reference-based"
    
  )

abundance_table_karst <- read_tsv(abundance_table_karst_path)

abundance_table_karst_new <- read_tsv(abundance_table_karst_new_path) %>% 
  mutate(
    pipeline = "karst", 
    polishing = "karst"
    
  )

```

# tables

## taxonomy list

```{r}
species_list <- abundance_table_karst %>% 
  select(Species) %>% 
  pull()
```

## compare abundance

### parse karst

```{r}
abundance_table_karst_parsed <-
  abundance_table_karst %>% 
  rename(
    Genus = Species 
  )
```

### parse umi-pipeline-nf

```{r}

# There was a change in taxonimic naming after the karst publication and
# Lactobacillus fermentum is now called Limosilactobacillus fermentum
abundance_table_parsed <-
  abundance_table_umi_poa %>% 
  bind_rows(abundance_table_umi_reference) %>% 
  # bind_rows(abundance_table_karst_new) %>% 
  separate(tax, into = paste0("Rank", 1:7), sep = ";", fill = "right") %>%
  rename(Genus = Rank7) %>% 
  select(
    Genus, 
    total, 
    pipeline, 
    polishing)

genus_abundance <-
  abundance_table_parsed %>%
  mutate(
    Genus = if_else(str_detect(Genus, "Limosilactobacillus"), "Lactobacillus", Genus),
    Genus = if_else(Genus %in% species_list, Genus, "other")
  ) %>% 
  group_by(Genus, pipeline, polishing) %>%
  summarise(
    total_abundance = sum(total, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  group_by(pipeline, polishing) %>% 
  mutate(
    total_reads = sum(total_abundance),
    relative_abundance = total_abundance / total_reads * 100, 
  ) %>% 
  # filter(relative_abundance > 0.5) %>% # keep genera with more than 0.5% abundance
  ungroup() %>%
  select(Genus, relative_abundance, pipeline, polishing, total_abundance, total_reads)

genus_abundance_parsed <- 
  genus_abundance %>% 
  select(Genus, relative_abundance, pipeline, polishing) %>%
  pivot_wider(
    names_from = c(pipeline, polishing),
    values_from = c(relative_abundance)
  )
```

### comparison
```{r}
comparison <-
  abundance_table_karst_parsed %>% 
  full_join(genus_abundance, by = "Genus") %>% 
  mutate_if(
    is.numeric, ~replace(., is.na(.), 0)
  ) %>% 
  mutate(
    across(c(ONT_10.3_UMI, Theoretical, relative_abundance), round, 2)
  ) %>%  
  select(
    Genus, ONT_10.3_UMI, Theoretical, relative_abundance, pipeline, polishing, total_abundance, total_reads
  )

comparison_long <- 
  comparison %>%
  select(Genus, Theoretical, ONT_10.3_UMI, relative_abundance, pipeline, polishing) %>% 
  rename(
    "longread_umi (Karst et al.)" = ONT_10.3_UMI, 
    # "other" = relative_abundance,
  ) %>% 
  pivot_longer(cols = c(relative_abundance), names_to = "Sample", values_to = "Abundance")

# r_pipelines <- cor(comparison$relative_abundance_umi_pipeline_nf, comparison$ONT_10.3_UMI)
# rsquare_pipelines <- summary(lm(ONT_10.3_UMI~relative_abundance_umi_pipeline_nf, data=comparison))$r.squared

correlation_pipelines <- 
  comparison %>% 
  group_by(polishing) %>% 
  group_nest() %>%
  mutate(
    model = map(data, ~ lm(ONT_10.3_UMI ~ relative_abundance, data = .x)),
    r_squared = map_dbl(model, ~ summary(.x)$r.squared),
    r = map_dbl(data, ~ cor(.x$relative_abundance, .x$ONT_10.3_UMI))
  ) %>%
  select(polishing, r_squared, r)

# Comparison with expected data
# Should we sequence the zymo sampleset and show that we are better than karst by now? In case we are. 
# PCR might induce many artifacts for us as well. 

# cor(comparison$ONT_10.3_UMI, comparison$Theoretical)
# cor(comparison$relative_abundance_umi_pipeline_nf, comparison$Theoretical)

```
### Explain differences in read number between POA and Refrerence-based polishing
```{r}
abundance_table_umi_poa %>%
  summarise(total_reads = sum(total))

abundance_table_umi_reference %>%
  summarise(total_reads = sum(total))

```

## Table

#### correlation
```{r}
# Round values for publication
correlation_pipelines <- correlation_pipelines %>%
  mutate(
    r_squared = round(r_squared, 3),
    r = round(r, 3)
  )

# Create flextable
ft_correlation_pipelines <- flextable(correlation_pipelines) %>%
  set_header_labels(
    polishing = "Polishing Method",
    r_squared = "R²",
    r = "Pearson's r"
  ) %>%
  align(align = "center", part = "all") %>%
  bold(part = "header") %>%
  fontsize(size = 11, part = "all") %>%
  padding(padding = 5, part = "all") %>% 
  theme_zebra() %>%
  border_remove() %>%
  border_outer(part = "all", border = fp_border(color = "black", width = 1)) %>%  
  border_inner_v(border = fp_border(color = "grey60", width = 0.5))%>%
  border_inner_h(border = fp_border(color = "grey60", width = 0.5)) 

# Export to Word
doc_correlation_pipelines <- read_docx() %>%
  body_add_par("Correlation Summary by Polishing Method", style = "heading 1") %>%
  body_add_flextable(ft_correlation_pipelines) %>%
  body_add_par("", style = "Normal")  # optional spacer

print(doc_correlation_pipelines, target = "Karst/out/SuppTable_correlation_pipelines.docx")
```

#### comparison
```{r}
# Create formatted flextable
ft_comparison <- flextable(comparison) %>%
  set_header_labels(
    Genus = "Genus",
    ONT_10.3_UMI = "longread_umi (Karst et al.)",
    # Theoretical = "Theoretical (%)",
    relative_abundance = "Relative Abundance (%)",
    pipeline = "Pipeline",
    polishing = "Polishing",
    total_abundance = "Total Abundance",
    total_reads = "Total Reads"
  ) %>%
  align(align = "center", part = "all") %>%
  bold(part = "header") %>%
  fontsize(size = 11, part = "all") %>%
  # padding(padding = 5, part = "all") %>% 
  theme_zebra() %>%
  border_remove() %>%
  border_outer(part = "all", border = fp_border(color = "black", width = 1)) %>%  
  border_inner_v(border = fp_border(color = "grey60", width = 0.5)) %>%
  border_inner_h(border = fp_border(color = "grey60", width = 0.5))

# Export to Word document
doc_comparison <- read_docx() %>%
  body_add_par("Relative Abundance Table by Genus", style = "heading 1") %>%
  body_add_flextable(ft_comparison)

print(doc_comparison, target = "Karst/out/SuppTable_comparison_abundance.docx")
```

## performance measures (precision, recall, species-level correlation)
Use the number of calssifiable reads for both umi-pipeline-nf polishing methods to definer precision and recall
use the relative abundances of karst, both umi-pipeline-nf polishing methods and theoretical to calculate the correlation at species level
```{r}
# total reads
abundance_table_umi_poa %>%
  summarise(total_reads = sum(total))
abundance_table_umi_reference %>%
  summarise(total_reads = sum(total))

# Use reads correctly assigned read counts to a genus to calculate precision and recall
performance_measures <-
genus_abundance %>%
  mutate(
    correctly_assigned = Genus != "other",
  ) %>%
  group_by(pipeline, polishing) %>%
  summarise(
    true_positives = sum(if_else(correctly_assigned, total_abundance, 0)),
    false_positives = sum(if_else(!correctly_assigned, total_abundance, 0)),
    false_negatives = sum(
      case_when(
        pipeline == "umi-pipeline-nf" & polishing == "POA-based" ~
          abundance_table_karst_parsed %>%
          filter(!Genus %in% genus_abundance$Genus[genus_abundance$pipeline == "umi-pipeline-nf" & genus_abundance$polishing == "POA-based"]) %>%
          pull(Genus) %>%
          length(),
        pipeline == "umi-pipeline-nf" & polishing == "Reference-based" ~
          abundance_table_karst_parsed %>%
          filter(!Genus %in% genus_abundance$Genus[genus_abundance$pipeline == "umi-pipeline-nf" & genus_abundance$polishing == "Reference-based"]) %>%
          pull(Genus) %>%
          length(),
        TRUE ~ 0
      )
    ),
    .groups = "drop"
  ) %>%
  mutate(
    precision = true_positives / (true_positives + false_positives),
    recall = true_positives / (true_positives + false_negatives), 
    f1_score = 2 * (precision * recall) / (precision + recall), 
  )
```

### create flextable for performance measures
```{r}
ft_performance_measures <- 
  performance_measures %>%
  select(pipeline, polishing, precision, recall, f1_score) %>%
  flextable() %>% 
  set_header_labels(
    pipeline = "Pipeline",
    polishing = "Polishing Method",
    precision = "Precision",
    recall = "Recall",
    f1_score = "F1 Score"
  ) %>%
  align(align = "center", part = "all") %>%
  bold(part = "header") %>% 
  fontsize(size = 11, part = "all") %>%
  padding(padding = 5, part = "all") %>% 
  theme_zebra() %>%
  border_remove() %>%
  border_outer(part = "all", border = fp_border(color = "black", width = 1)) %>%  
  border_inner_v(border = fp_border(color = "grey60", width = 0.5))%>%
  border_inner_h(border = fp_border(color = "grey60", width = 0.5))

# Export to Word
doc_performance_measures <- read_docx() %>%
  body_add_par("Performance Measures by Polishing Method", style = "heading 1") %>%
  body_add_flextable(ft_performance_measures) %>%
  body_add_par("", style = "Normal")  # optional spacer 

print(doc_performance_measures, target = "Karst/out/SuppTable_performance_measures.docx")
```


### correlation
#### POA-based
```{r}
corraltion_poa <- 
comparison %>% 
  filter(polishing == "POA-based") %>%
  summarise(
    correlation_karst = cor(ONT_10.3_UMI, relative_abundance), 
    correlation_ground_truth = cor(Theoretical, relative_abundance)
  ) %>% 
  mutate(
    method = "umi-pipeline-nf (POA-based)"
  )
```

#### Reference-based

```{r}
correlation_reference <-
comparison %>% 
  filter(polishing == "Reference-based") %>%
  summarise(
    correlation_karst = cor(ONT_10.3_UMI, relative_abundance), 
    correlation_ground_truth = cor(Theoretical, relative_abundance)
  ) %>% 
  mutate(
    method = "umi-pipeline-nf (Reference-based)"
  )

```

#### Karst

```{r}
correlation_karst <-
comparison %>% 
  summarise(
    correlation_ground_truth = cor(ONT_10.3_UMI, Theoretical)
  ) %>% 
  mutate(
    method = "longread_umi (Karst et al.)"
  )
```

#### combine

```{r}
correlation_summary <- 
  bind_rows(corraltion_poa, correlation_reference, correlation_karst)
  
```

### create flextable
```{r}
ft_correlation_summary <- 
  correlation_summary %>% 
  select(method, correlation_karst, correlation_ground_truth) %>%
  flextable() %>% 
  set_header_labels(
    method = "Method",
    correlation_karst = "Correlation with Karst et al.",
    correlation_ground_truth = "Correlation with Theoretical"
  ) %>%
  align(align = "center", part = "all") %>%
  bold(part = "header") %>%
  fontsize(size = 11, part = "all") %>%
  padding(padding = 5, part = "all") %>% 
  theme_zebra() %>%
  border_remove() %>%
  border_outer(part = "all", border = fp_border(color = "black", width = 1)) %>%  
  border_inner_v(border = fp_border(color = "grey60", width = 0.5))%>%
  border_inner_h(border = fp_border(color = "grey60", width = 0.5))

# Export to Word
doc_correlation_summary <- read_docx() %>%
  body_add_par("Correlation Summary", style = "heading 1") %>%
  body_add_flextable(ft_correlation_summary) %>%
  body_add_par("", style = "Normal")  # optional spacer

print(doc_correlation_summary, target = "Karst/out/SuppTable_correlation_summary.docx")


```


# Plots
## compare karst vs umi-pipeline-nf vs theoretical abundance

```{r}

p.relative_abundance_comparison_polishing <-
  comparison %>% 
  ggplot(aes(x = ONT_10.3_UMI, y = relative_abundance)) +
  geom_smooth(method = "lm", color = "black", fill = "grey60") +
  geom_point(aes(color = Genus), size = 3) +
  facet_wrap(vars(polishing)) + 
  labs(
    x = "longread_umi (Karst et al.)", 
    y = "umi-pipeline-nf", 
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 22, color = "black"), 
    axis.text = element_text(colour = "black")
  )

ggsave(
  plot = p.relative_abundance_comparison_polishing,
  filename = "relative_abundance_comparison_polishing.jpeg", 
  path = "Karst/out/",
  width = 12, 
  height = 9
)
```

# Create a stacked barpolot to compare the abundance profiles for Karst, Theoretical and umi-pipeline-nf both methods

```{r}
plot_naming <- 
  c(
    "umi-pipeline-nf_POA-based" = "umi-pipeline-nf\n(POA-based)", 
    "umi-pipeline-nf_Reference-based" = "umi-pipeline-nf\n(Reference-based)", 
    "ONT_10.3_UMI" = "longread_umi\n(Karst et al.)",
    # "ONT_meta" = "ONT meta\n(Karst et al.)",
    "Theoretical" = "Theoretical\n(Zymo Research)"
  )
plot_order <- 
  c(
    "umi-pipeline-nf_Reference-based",
    "umi-pipeline-nf_POA-based", 
    "ONT_10.3_UMI",
    "Theoretical"
    # "ONT_meta",
  )

plot_colors <- 
  c(
    "Bacillus" = "#1f77b4",
    "Escherichia" = "#ff7f0e",
    "Enterococcus" = "#ffbb78",
    "Lactobacillus" = "#2ca02c",
    "Listeria" = "#d62728",
    "Pseudomonas" = "#9467bd",
    "Salmonella" = "#8c564b",
    "Staphylococcus" = "#e377c2",
    "other" = "#7f7f7f"
  )

plot_colors <- c(
  "Bacillus"     = "#66c2a5",
  "Escherichia"  = "#fc8d62",
  "Enterococcus" = "#8da0cb",
  "Lactobacillus"= "#a6d854",
  "Listeria"     = "#e78ac3",
  "Pseudomonas"  = "#ffd92f",
  "Salmonella"   = "#e5c494",
  "Staphylococcus" = "#80b1d3",  # ergänzend, dezentes Blau
  "other"        = "#8c8c8c"     # neutral grau
)

plot_colors <- c(
  "Bacillus"     = "#6D90A6",
  "Escherichia"  = "#A6BD8A",
  "Enterococcus" = "#C9A3A9",
  "Lactobacillus"= "#8DB0AF",
  "Listeria"     = "#B9A56A",
  "Pseudomonas"  = "#8D7F8D",
  "Salmonella"   = "#A2C4A2",
  "Staphylococcus" = "#9CA3B3",
  "other"        = "#7F7F7F"
)

# caluclate the least difference of each genus of umi-pipeline-nf_Reference-based and Theoretical and order the colors accordingly
genus_order <- 
  comparison %>%
  filter(polishing == "Reference-based") %>%
  mutate( 
    difference = abs(relative_abundance - Theoretical)
  ) %>% 
  arrange(desc(difference)) %>% 
  pull(Genus)

genus_order <- 
  comparison %>%
  filter(polishing == "Reference-based") %>%
  arrange(desc(relative_abundance)) %>% 
  pull(Genus)

comparison_barplot <- 
  genus_abundance_parsed %>% 
  full_join(abundance_table_karst_parsed, by = "Genus") %>% 
  select(
    -PacBio_UMI, -PacBio_CCS, -ONT_meta
  ) %>% 
  pivot_longer(
    cols = c(Theoretical, ONT_10.3_UMI, `umi-pipeline-nf_POA-based`, `umi-pipeline-nf_Reference-based`), names_to = "Pipeline", values_to = "Abundance"
  ) %>% 
  mutate(
    Pipeline = factor(Pipeline, levels = plot_order), 
    Genus = factor(Genus, levels = genus_order),
    Abundance = replace_na(Abundance, 0)
  )

comparison_barplot %>% 
  write_tsv("main_figure/stacked_barplot_comparison_theoretical.tsv")


p.stacked_barplot_theoretical <-
  comparison_barplot %>%
  ggplot(aes(x = Pipeline, y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = plot_colors) +
  scale_x_discrete(labels = plot_naming) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 102), breaks = seq(0, 100, 10)) +
  labs(
    x = "", 
    y = "Relative abundance [%]",
    # title = "Genus-Level Comparison of Abundances"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    text = element_text(size = 22, colour = "black"),
    axis.text = element_text(colour = "black"),
    # axis.text.x = element_text(angle = 45, hjust = 1)
  )

p.stacked_barplot_theoretical

ggsave(
  plot = p.stacked_barplot_theoretical,
  filename = "stacked_barplot_comparison_theoretical.jpeg", 
  path = "Karst/out/",
  width = 12, 
  height = 9
)

```

# Lunch seminar

## correlation

```{r}
comparison %>% 
  ggplot(aes(x = ONT_10.3_UMI, y = relative_abundance)) +
  geom_smooth(method = "lm", color = "black", fill = "grey60") +
  geom_point(aes(color = Genus), size = 3) +
  facet_wrap(vars(polishing)) + 
  labs(
    x = "longread_umi (Karst et al.)", 
    y = "umi-pipeline-nf", 
    # title = "Relative abundance of each Genus (ONT MinION R10.3)"
  ) +
  theme_bw() +
  theme(
    text = element_text(size = 28, color = "black"), 
    axis.text = element_text(colour = "black")
  )

ggsave(
  filename = "lunchseminar_benchmarking_karst.jpeg", 
  path = "lunch_seminar/out/",
  dpi = 600,
  width = 12, 
  height = 9
)
```
