---
title: "main_figure_new"
format: html
editor: visual
---
# Libraries

```{r}
library(tidyverse)
library(patchwork)
library(jpeg)
library(grid)
```

# Paths

```{r}
zymo_comparison_path <- "main_figure/data/stacked_barplot_comparison_theoretical.tsv"
umi_pipeline_nf_metropmap_path <- "main_figure/data/metro-map_umi-pipeline-nf.jpg"
scalability_runtime_path <- "main_figure/data/scalability_plot.tsv"
scatter_variant_level_data_path <- "main_figure/data/scatter_variant_level_data.tsv"
early_stopping_cluster_stats_parsed_path <- "main_figure/data/early_stopping_cluster_stats_parsed.tsv"
early_stopping_variant_calling_comparison_path <- "main_figure/data/early_stopping_variant_calling_comparison.tsv"
full_run_cluster_stats_parsed_path <- "main_figure/data/full_run_cluster_stats_parsed.tsv"
```

# Read data

```{r, message=FALSE}
zymo_comparison <- read_tsv(zymo_comparison_path)
# umi_pipeline_nf_metropmap <- readJPEG(umi_pipeline_nf_metropmap_path)
scalability_runtime <- read_tsv(scalability_runtime_path) 
scatter_variant_level_data <- read_tsv(scatter_variant_level_data_path)
early_stopping_cluster_stats_parsed <- read_tsv(early_stopping_cluster_stats_parsed_path) %>% 
  mutate(mode = "early")
early_stopping_variant_calling_comparison <- read_tsv(early_stopping_variant_calling_comparison_path)
full_run_cluster_stats_parsed <- read_tsv(full_run_cluster_stats_parsed_path) %>% 
  mutate(mode = "full")
```
# Global sizing and theme

```{r}
text_size <- 16
annotation_size <- 5
```

## plot
```{r}
samples <- unique(full_run_cluster_stats_parsed$sample)
# "HG01593" "HG02657" "HG02778" "NA19098" "NA19190" "HG03765" "HG03234" "NA18868"
clustering_colors <- 
  c(
    "HG01593" = "#1b9e77",
    "HG02657" = "#d95f02",
    "HG02778" = "#7570b3",
    "NA19098" = "#e7298a",
    "NA19190" = "#66a61e",
    "HG03765" = "#e6ab02",
    "HG03234" = "#a6761d",
    "NA18868" = "#666666"
  )

p.cluster_stats_parsed <-
  full_run_cluster_stats_parsed %>%
  ggplot(aes(
    x = runtime_hours,
    y = n_cluster,
    color = sample,
  )) +
  geom_smooth(se = FALSE) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
  scale_color_manual(values = clustering_colors) +
  labs(
    # title = "Number of clusters per sample over time",
    x = "Runtime [h]",
    y = "UMI-Clusters [n]",
    color = "Sample"
  ) +
  theme_bw() +
  theme(
    legend.position = "left",
    text = element_text(size = text_size)
  )

p.cluster_stats_parsed
```
# variant calling comparison early stopping
```{r}
variant_calling_correlation <- 0.999
n_variants <- nrow(early_stopping_variant_calling_comparison)

p.variant_calling_early_stopping <-
  early_stopping_variant_calling_comparison %>%
  ggplot(aes(x = variant_level_full, y = variant_level_early)) +
  geom_point(size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey80") +
  geom_text(
    aes(
      x = 0.13,
      y = 0.98,
      # paste R-saquared value with 4 decimal places
      label = paste0("R^2 == ", variant_calling_correlation)
    ),
    parse = TRUE,
    size = annotation_size
  ) +
  # add number of variants per condition
  geom_text(
    aes(
      x = 0.24,
      y = 0.85,
      label = paste0("Variants = ", n_variants, " [n]")
    ),
    size = annotation_size
  ) +
  # make percentage scale
  scale_x_continuous(labels = scales::percent, breaks = seq(0, 1, 0.2)) +
  scale_y_continuous(labels = scales::percent, breaks = seq(0, 1, 0.2)) +
  labs(
    # title = "Variant Level Correlation between Early Stopping and Full Run",
    x = "Variant Level (Full Run)",
    y = "Variant Level\n(Early Stopping)"
  ) +
  theme_bw() +
  theme(
    text = element_text(size = text_size)
  )

p.variant_calling_early_stopping
```

# Scalability
## both faceted
```{r}
colors <- 
  c(
    "CPU" = "#E64B35",
    "GPU" = "#4DBBD5"
  )

p.scalability_runtime <- 
  scalability_runtime %>%
  mutate(
    total_runtime_hours = total_runtime_sec / 3600,
  ) %>% 
  ggplot(aes(
    x = sample_num,
    y = total_runtime_hours,
    color = mode,
    shape = type
  )) +
  facet_wrap(~ polishing, nrow = 1, scales = "free_y") +
  geom_point(size = 2) +
  geom_line(aes(linetype = type)) +
  scale_y_continuous(breaks = scales::breaks_pretty(n = 8)) +
  # scale_x_continuous(breaks = seq(0, 100, 20)) +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 6)) +
  labs(
    x = "Samples [n]",
    y = "Total Runtime [h]",
    color = "Mode",
    shape = "Data type",
    linetype = "Data type"
  ) +
  scale_color_manual(values = colors) +
  theme_bw() +
  theme(
    legend.position = "left",
    text = element_text(size = text_size), 
    # increase facet label size 
    strip.text = element_text(size = text_size)
  )

p.scalability_runtime
```
## POA-based

```{r}
colors <- 
  c(
    "CPU" = "#E64B35",
    "GPU" = "#4DBBD5"
  )

p.scalability_runtime_poa <- 
  scalability_runtime %>%
  filter(polishing == "POA-based") %>%
  mutate(
    total_runtime_hours = total_runtime_sec / 3600,
  ) %>% 
  ggplot(aes(
    x = sample_num,
    y = total_runtime_hours,
    color = mode,
    shape = type
  )) +
  geom_point(size = 2) +
  geom_line(aes(linetype = type)) +
  scale_y_continuous(breaks = scales::breaks_pretty(n = 8)) +
  # scale_x_continuous(breaks = seq(0, 100, 20)) +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 6)) +
  labs(
    x = "Samples [n]",
    y = "Total Runtime [h]",
    color = "Mode",
    shape = "Data type",
    linetype = "Data type"
  ) +
  scale_color_manual(values = colors) +
  theme_bw() +
  theme(
    legend.position = "left",
    text = element_text(size = text_size), 
    # increase facet label size 
    strip.text = element_text(size = text_size)
  )
```

## Reference-based

```{r}
colors <- 
  c(
    "CPU" = "#E64B35",
    "GPU" = "#4DBBD5"
  )

p.scalability_runtime_reference <- 
  scalability_runtime %>%
  filter(polishing == "Reference-based") %>%
  mutate(
    total_runtime_hours = total_runtime_sec / 3600,
  ) %>% 
  ggplot(aes(
    x = sample_num,
    y = total_runtime_hours,
    color = mode,
    shape = type
  )) +
  geom_point(size = 2, show.legend = FALSE) +
  geom_line(aes(linetype = type), show.legend = FALSE) +
  scale_y_continuous(breaks = scales::breaks_pretty(n = 8)) +
  # scale_x_continuous(breaks = seq(0, 100, 20)) +
  scale_x_continuous(breaks = scales::breaks_pretty(n = 6)) +
  labs(
    x = "Samples [n]",
    y = "Total Runtime [h]",
    color = "Mode",
    shape = "Data type",
    linetype = "Data type"
  ) +
  scale_color_manual(values = colors) +
  theme_bw() +
  theme(
    legend.position = "left",
    text = element_text(size = text_size), 
    # increase facet label size 
    strip.text = element_text(size = text_size)
  )

p.scalability_runtime_reference
```

# Zymo comparison

```{r}
plot_naming <- 
  c(
    "umi-pipeline-nf_POA-based" = "umi-pipeline-nf\n(POA-based)", 
    "umi-pipeline-nf_Reference-based" = "umi-pipeline-nf\n(Reference-based)", 
    "ONT_10.3_UMI" = "longread_umi\n(Karst et al.)",
    "Theoretical" = "Theoretical\n(Zymo Research)"
  )

plot_order <- 
  c(
    "umi-pipeline-nf_Reference-based",
    "umi-pipeline-nf_POA-based", 
    "ONT_10.3_UMI",
    "Theoretical"
  )
# similar colors as in cluster monitoring experiment
plot_colors <- 
  c(
    "Staphylococcus" = "#1b9e77",
    "Lactobacillus" = "#d95f02",
    "Bacillus" = "#7570b3",
    "Listeria" = "#e7298a",
    "Escherichia" = "#66a61e",
    "Pseudomonas" = "#e6ab02",
    "Salmonella" = "#a6761d",
    "Enterococcus" = "#666666",
    "other" = "#999999"
  )
  
# genus order 
genus_order <- 
  c(
    "Staphylococcus",
    "Lactobacillus",
    "Bacillus",
    "Listeria",
    "Escherichia",
    "Pseudomonas",
    "Salmonella",
    "Enterococcus",
    "other"
  )

p.stacked_barplot_theoretical <-
  zymo_comparison %>%
  mutate(
    Pipeline = factor(Pipeline, levels = plot_order), 
    Genus = factor(Genus, levels = genus_order),
  ) %>%
  ggplot(aes(x = Pipeline, y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity", width = 0.8) +
  scale_fill_manual(values = plot_colors) +
  scale_x_discrete(labels = plot_naming) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 102), breaks = seq(0, 100, 20)) +
  labs(
    x = "", 
    y = "Relative\nAbundance [%]"
  ) +
  theme_bw() +
  theme(
    legend.position = "left"
  )

p.stacked_barplot_theoretical
```



# Combine plots scalability, cluster monitoring, variant calling comparison 
```{r}
design_figure_main <- "AA\nBC\nDD"

pw.main_figure <- 
  wrap_plots( 
    A = free(p.scalability_runtime),
    B = free(p.cluster_stats_parsed),
    C = free(p.variant_calling_early_stopping),
    D = free(p.stacked_barplot_theoretical),
    design = design_figure_main,
    widths = c(7, 6),
    heights = c(4, 4, 4)
  ) + 
  plot_annotation(tag_levels = "A") &
  theme(
    text = element_text(size = text_size),
    axis.text = element_text(size = text_size, colour = "black")
    # legend.position = "bottom",
    # plot.tag = element_text(size =, face = "bold")
  )

pw.main_figure

ggsave(
  filename = "main_figure/out/main_figure.jpg",
  plot = pw.main_figure,
  dpi = 600,
  width = 12,
  height = 12
)

```



# Combine plots scalability, cluster monitoring, variant calling comparison 
```{r}
design_figure_main_2 <- "AB\nCD\nEE"

pw.main_figure_2 <- 
  wrap_plots( 
    A = free(p.scalability_runtime_poa),
    B = free(p.scalability_runtime_reference),
    C = free(p.cluster_stats_parsed),
    D = free(p.variant_calling_early_stopping),
    E = free(p.stacked_barplot_theoretical),
    design = design_figure_main_2
    # widths = c(6, 6)
    # heights = c(4, 4, 4)
  ) + 
  plot_annotation(tag_levels = "A") &
  theme(
    text = element_text(size = text_size),
    axis.text = element_text(size = text_size, colour = "black")
    # legend.position = "bottom",
    # plot.tag = element_text(size =, face = "bold")
  )

pw.main_figure_2

ggsave(
  filename = "main_figure/out/main_figure_2.jpg",
  plot = pw.main_figure_2,
  dpi = 600,
  width = 12,
  height = 12
)
```


# Combine plots scalability, cluster monitoring, variant calling comparison with shared axes for scalability
```{r}
# Sub-assembly for A and B with shared axes (and a single set of axis titles)
top_AB <- 
  wrap_plots(
    A = p.scalability_runtime_poa,
    B = p.scalability_runtime_reference, 
    design = "AB"
  ) + 
  plot_layout(
    axes = "collect", 
    axis_titles = "collect"
  )

# Use the sub-assembly as a single block in the main design
design_figure_main_3 <- "AA\nCD\nEE"

pw.main_figure_3 <- 
  wrap_plots( 
    A = top_AB,
    C = free(p.cluster_stats_parsed),
    # C = p.cluster_stats_parsed,
    D = free(p.variant_calling_early_stopping),
    # D = p.variant_calling_early_stopping,
    E = free(p.stacked_barplot_theoretical),
    # E = p.stacked_barplot_theoretical,
    design = design_figure_main_3, 
    widths = c(6, 8),
    heights = c(4, 4, 4)
  ) + 
  plot_annotation(tag_levels = "A") &
  theme(
    text = element_text(size = text_size),
    axis.text = element_text(size = text_size, colour = "black")
  )

pw.main_figure_3

ggsave(
  filename = "main_figure/out/main_figure_3.jpg",
  plot = pw.main_figure_3,
  dpi = 600,
  width = 12,
  height = 12
)
```


