---
title: "SARS-CoV-2_benchmarking"
format: html
editor: visual
---

```{r setup, include=FALSE}
library(tidyverse)
library(Biostrings)
library(networkD3)
library(flextable)
library(scales)
library(officer)
library(patchwork)
```

# functions

```{r}
make_publication_table <- function(df, caption = NULL) {
  flextable(df) %>%
    autofit() %>%
    {if (!is.null(caption)) set_caption(., caption) else .} %>%
    align(align = "center", part = "all") %>%
    bold(part = "header") %>%
    fontsize(size = 11, part = "all") %>%
    padding(padding = 5, part = "all") %>%
    theme_zebra() %>%
    border_remove() %>%
    border_outer(part = "all", border = fp_border(color = "black", width = 1)) %>%
    border_inner_v(border = fp_border(color = "grey60", width = 0.5)) %>%
    border_inner_h(border = fp_border(color = "grey60", width = 0.5))
}
```

# Paths

```{r}
nextclade_paths <- list.files(
  path = "../ConSeqUmi/output/", 
  pattern = "nextclade.tsv", 
  recursive = TRUE, 
  full.names = TRUE
)

cluster_stats_paths <- list.files(
  path = "../ConSeqUmi/cluster_stats/", 
  pattern = "cluster_stats.tsv", 
  recursive = TRUE, 
  full.names = TRUE
)
sample_sheet_path <- "../ConSeqUmi/sample_sheet.tsv" 
```

# load files

```{r}
nextclade <- lapply(nextclade_paths, function(path){
  barcode <- str_extract(path, "barcode\\d\\d")
  
  read_tsv(path, show_col_types = FALSE) %>% 
    mutate(
      barcode = barcode
    )
}) %>% 
  bind_rows()

cluster_stats <- lapply(cluster_stats_paths, function(path){
  barcode <- str_extract(path, "barcode\\d\\d")
  
  read_tsv(path, show_col_types = FALSE) %>% 
    mutate(
      barcode = barcode
    )
}) %>% 
  bind_rows()

sample_sheet <- read_tsv(sample_sheet_path)
```

# filter cluster stats

```{r}
min_reads_per_cluster <- 12

cluster_stats_filtered <- 
  cluster_stats %>% 
  filter(cluster_written == 1) %>% 
  filter(reads_found >= min_reads_per_cluster) %>% 
  mutate(
    reads_found = reads_found_fwd + reads_found_rev
  ) %>% 
  select(cluster_id, barcode, reads_found)

```

# Filter Nextclade results

Remove replicate of sample#1. Too much for the little bechmark in the paper and data of an R10 flongle might anyway not be very reliable. Filter for Spike variants only

```{r}
nextclade_filtered <- 
  nextclade %>% 
  mutate(
    aaSubstitutions = aaSubstitutions %>%
      str_split(",") %>%
      map(~ .x[str_detect(.x, "^S:")]) %>%
      map_chr(~ if (length(.x) > 0) paste(.x, collapse = ",") else NA_character_)
  ) %>% 
  drop_na(aaSubstitutions) %>% 
  group_by(barcode, aaSubstitutions) %>% 
  mutate(n = n()) %>% 
  filter(str_detect(barcode, "barcode04", negate = TRUE)) %>%
  left_join(cluster_stats_filtered, by = c("seqName" = "cluster_id", "barcode")) %>% 
  select(
    seqName,
    clade, 
    clade_display, 
    clade_nextstrain, 
    Nextclade_pango, 
    qc.overallStatus,
    substitutions, 
    aaSubstitutions, 
    aaDeletions, 
    aaInsertions, 
    `founderMuts['clade'].nodeName`, 
    `founderMuts['clade'].aaSubstitutions`, 
    `founderMuts['Nextclade_pango'].aaDeletions`, 
    barcode, 
    reads_found, 
    n
  ) %>% 
  inner_join(sample_sheet)
```

# Variant classification

```{r}
variant_classification <- 
  nextclade_filtered %>% 
  group_by(clade_display, sample) %>% 
  summarise(
    n = n(), 
    .groups = "drop"
  ) %>% 
  group_by(sample)%>% 
  mutate(
    total = sum(n), 
    percent = n / total * 100
  )
  
variant_classification
```

## Table

```{r}
table_variant_classification <- 
  variant_classification %>% 
  mutate(
    percent = round(percent, 2)
  ) %>%
  arrange(sample) %>% 
  select(
   Sample = sample, 
   "Clade (Strain)" = clade_display, 
   "Consensus sequences per clade [n]" = n, 
   "Consenus sequences per clade [%]" = percent,
   "Total consensus sequences per sample [n]" = total, 
  )

publication_table_variant_classification <- 
  make_publication_table(table_variant_classification, "SARS-CoV-2 viral lineage assignment (Nextclade)")
  
doc_variant_classification <- 
  read_docx() %>%
  body_add_flextable(publication_table_variant_classification)

print(doc_variant_classification, target = "../ConSeqUmi/out/SuppTable_doc_variant_classification.docx")
```

# Diversity

```{r}
diversity <- 
  nextclade_filtered %>% 
  group_by(sample) %>%
  mutate(
    total = n()
  ) %>% 
  group_by(sample, aaSubstitutions, total) %>% 
  summarise(
    n = n()
  )

unique_spikes <- 
  diversity %>% 
  select(sample, aaSubstitutions, total) %>% 
  group_by(sample, total) %>% 
  summarise(
    n = n()
  )

```

# clades

```{r}
nextclade_filtered %>% 
  group_by(sample, clade) %>% 
  summarise(n = n())

```

# Patient #1

```{r}
patient1 <- 
  nextclade_filtered %>% 
  filter(sample == "patient1") %>% 
  filter(replicate == "replicate1") %>% 
  mutate(
    A222V = str_detect(aaSubstitutions, "A222V"), 
    V367L = str_detect(aaSubstitutions, "V367L"), 
    Y473 = str_detect(aaSubstitutions, "473"), 
    Y145H = str_detect(aaSubstitutions, "Y145H"), 
    I834V = str_detect(aaSubstitutions, "I834V"), 
    I834 = str_detect(aaSubstitutions, "I834"), 
    I882 = str_detect(aaSubstitutions, "882"), 
    G142D = str_detect(aaSubstitutions, "G142D"), 
  )
```

## Unique spikes

We found 170 unique spike regions in 702 consensus sequences

```{r}
patient1 %>% 
  nrow()

patient1 %>% 
  group_by(aaSubstitutions) %>% 
  summarise(n = n())

patient1 %>% 
  select(clade, aaSubstitutions) %>% 
  group_by(clade, aaSubstitutions) %>% 
  summarise(count = n()) %>% 
  group_by(clade) %>% 
  summarise(
    unique = sum(count), 
    n = n()
  )

```

## 21A/J vs 21I

```{r}
subclades <- 
patient1 %>% 
  group_by(clade_display) %>% 
  summarise(
    n = n(), 
    .groups = "drop"
  ) %>% 
  mutate(
    total = sum(n), 
    percent = n / total * 100
  )
```

### Table

```{r}
table_subclades <- 
  subclades %>% 
  mutate(
    percent = round(percent, 2)
  ) %>%
  arrange(percent) %>% 
  select(
   "Clade (Strain)" = clade_display, 
   "Consensus sequences per clade [n]" = n, 
   "Consenus sequences per clade [%]" = percent,
   "Total consensus sequences [n]" = total, 
  )

publication_table_subclades <- 
  make_publication_table(table_subclades, "SARS-CoV-2 viral lineage assignment of patient 1 (Nextclade)")
  
doc_subclades <- 
  read_docx() %>%
  body_add_flextable(publication_table_subclades)

print(doc_subclades, target = "../ConSeqUmi/out/SuppTable_doc_subclades.docx")
```

## diverse mutations

About half of the 21I variants carried both mutations. We could not find this mutations in about the other half of the variants.

```{r}
patient1 %>%
  group_by(clade, Nextclade_pango, A222V, V367L, Y473, Y145H, G142D, I834V) %>% 
  summarise(n = n())
```

## A222V and V367L

About half of the 21I variants carried both mutations. We could not find this mutations in about the other half of the variants.

```{r}
subclade_specific_variants <- 
patient1 %>%
  group_by(clade_display, A222V, V367L, G142D, I834V, I882) %>% 
  summarise(
    n = n(),
    .groups = "drop"
  ) %>% 
  mutate(
    total = sum(n), 
    percent = n / total * 100
  )
```

### Table

```{r}
table_subclade_specific_variants <- 
  subclade_specific_variants %>% 
  arrange(clade_display, percent) %>% 
  mutate(
    percent = round(percent, 2)
  ) %>%
  arrange(percent) %>% 
  select(
   "Clade (Strain)" = clade_display, 
   "Carries A222V" = A222V, 
   "Carries V367L" = V367L, 
   "Carries G142D" = G142D, 
   "Carries I834V" = I834V,
   "Carries I882" = I882,
   "Consenus sequences per clade [n]" = n,
   "Consenus sequences per clade [%]" = percent,
   "Total consensus sequences [n]" = total, 
  )

publication_table_subclade_specific_variants <- 
  make_publication_table(table_subclade_specific_variants, "SARS-CoV-2 viral lineage assignment of patient 1 (Nextclade)")
  
doc_subclade_specific_variants <- 
  read_docx() %>%
  body_add_flextable(publication_table_subclade_specific_variants)

print(doc_subclade_specific_variants, target = "../ConSeqUmi/out/SuppTable_doc_subclade_specific_variants.docx")
```

### 21A/J clade not containing G142D

There are publications reporting back mutation of G142D to the WT allele Spike Protein NTD mutation G142D in SARS-CoV-2 Delta VOC lineages is associated with frequent back mutations, increased viral loads, and immune evasion

```{r}
patient1 %>% 
  filter(!G142D)
```

## Lineages Pango

```{r}
patient1 %>% 
  group_by(Nextclade_pango) %>% 
  summarise(n = n())
```

## G142D in Delta variants

Usually ussed to classify delta variants - but some A/J variants don't contain this mutation

```{r}
patient1 %>% 
  group_by(clade, G142D) %>% 
  summarise(n = n())
```

## A222V and 882 carriers

We did not find any strains that contained the I882 ATC mutation.

```{r}
patient1 %>% 
  filter(A222V & I882)

patient1 %>% 
  filter(I882)
```

## I834V carriers

We can confirm this finding as well and also online this mutation is seen in AY.103 lineages

```{r}
patient1 %>% 
  group_by(Nextclade_pango, I834V) %>% 
  summarise(
    n = n(), .groups = "drop") %>% 
  mutate(
    total = sum(n),
    percentage = n / total * 100
  )
```
